---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# oaire

<!-- badges: start -->
<!-- badges: end -->

The `oaire` package offers an interface between R and the [OpenAIRE Graph API version 3](https://graph.openaire.eu/docs/apis/graph-api/). The package is built around four main functions:

-   `oag_query()` to build a valid query
-   `oag_options()` to set the options of a query (sorting, paging...)
-   `oag_request()` to send a request to the OpenAIRE Graph API
-   `oag_fetch()` to wrap in a single function building a query, making a request, and getting the data from the API

The most straightforward approach is to use `oag_fetch()` combined with `oag_options()`. Another approach is to make your own workflow: 1) building a query with `oa_query()`, 2) specifying options with `oag_options()` if necessary, 3) sending a request to the API with `oag_request()`.

The advantage of using `oag_fetch()` is that it handles paging automatically. `oag_query()` combined with `oag_request()` is better suited to retrieve a single page and/or get the number of records associated to a specific query.

## Installation

You can install the development version of `oaire` like so:

``` r
devtools::install("snsf-data/oaire")
```

## Authenticated request

The OpenAIRE Graph API can be accessed via authenticated and non-authenticated requests (see its [documentation on authenticated requests](https://graph.openaire.eu/docs/apis/authentication)). You can use a personal access token to make an authenticated request (granting 7'200 requests per hour). Two types of tokens exists: regular token and refresh token. The regular token is valid only one hour and must be renewed in your OpenAIRE Graph account after the validity is over. The refresh token is valid for one month and allows to generate access tokens that are valid for one hour. Access tokens can be renewed during the whole validity period of the refresh token.

If `oaire` cannot find any token in your environment variables, it will by default make an non-authenticated request (limited to 60 requests per hour). When a valid refresh token can be found in the environment variables, `oaire` will handle the refresh of the access token.

To make your token available to `oaire`, save it in your `.Renviron` file (it can be find at the user's home directory or it can be opened directly by running `usethis::edit_r_environ()`). To add a regular token, add the following line:

```
oag_api_token="<YOUR_REGULAR_TOKEN>"
```

To add a refresh token, add the following line:

```
oag_api_refresh_token="<YOUR_REFRESH_TOKEN>"
```

## Getting data from OpenAIRE Graph API

The OpenAIRE Graph API works around the idea of entities and filters. Entities have their own endpoints (see the [documentation of the OpenAIRE Graph API](https://graph.openaire.eu/docs/apis/home) for a list of the different entities and their respective filters).

In the following example, we query the "research-products" entity, requesting to return only the first record for the type "publication". By looking at the header of the server's response, we can know how many "publications" the database contains.

```{r}
library(oaire)

# Get the the first record of the research product type "publication"
n_pub <- oag_query(
  entity = "research-products",
  type = "publication",
  options = oag_options(page = 1, pageSize = 1)
) |>
  oag_request()

# Print the number of records available in the response's header
n_pub[["header"]][["numFound"]]
```

Maybe we are interested in dataset that were published on 15 January 2026:

```{r}
# Get the the first record of the research product type "publication", published
# on 15 January 2026.
n_pub_2026_15_01 <- oag_query(
  entity = "research-products",
  type = "dataset",
  fromPublicationDate = "2026-01-15",
  toPublicationDate = "2026-01-15",
  options = oag_options(page = 1, pageSize = 1)
) |>
  oag_request()

# Print the number of records available in the response's header
n_pub_2026_15_01[["header"]][["numFound"]]
```

The results from the last query shows that the number of records is reasonable enough to consider using paging to fetch in a single call all the data related to the datasets published on 15 January 2026. In the example below, we use offset-based paging, setting the number of records returned per page to 100 (the maximum allowed by the OpenAIRE Graph API): 

```{r}
# Fetch data for datasets published on 15 January 2026, sorted by descending
# date of publication, using offset-based-paging with 100 records per page.
datasets_2026_15_01_offset <- oag_fetch(
  entity = "research-products",
  type = "dataset",
  fromPublicationDate = "2026-01-15",
  toPublicationDate = "2026-01-15",
  options = oag_options(sortBy = c(publicationDate = "DESC"), pageSize = 100)
)
```

Offset-based paging is limited to small dataset of 10'000 records maximum, with the numbers of records being defined by the product of the `size` and `pageSize` options. When the dataset is larger than 10'000 records, cursor-based paging should be used. The example above can be reproduced using cursor-based paging by setting `cursor = TRUE` in `oag_options()`:

```{r}
# Fetch data for datasets published on 15 January 2026, sorted by descending
# date of publication, using cursor-based-paging with 100 records per page.
datasets_2026_15_01_cursor <- oag_fetch(
  entity = "research-products",
  type = "dataset",
  fromPublicationDate = "2026-01-15",
  toPublicationDate = "2026-01-15",
  options = oag_options(
    sortBy = c(publicationDate = "DESC"),
    pageSize = 100,
    cursor = TRUE
  )
)
```

More complex queries can be constructed using logical operators. For example, we could search for software in addition to datasets by combining the two types with an "OR" logical operator: `"dataset" OR "software"`. However, the OpenAIRE Graph API requires that values used with logical operators are wrapped with double quotes, which requires to escape them when creating the string passed to the "type" filter: `type = "\"dataset\" OR \"software\""`. To reduce the inconvenience, `oaire` provides a family of concatenation functions allowing to add "AND" (`concat_and()`), "OR" (`concat_or()`), and "NOT" (`concat_not()`) logical operators.

In the next example, we create a filter to find any "dataset" or "software" research product type, and use it to fetch data from the OpenAIRE Graph API.

```{r}
# Concatenate the strings "dataset" and "software" with "OR"
ds_and_soft <- concat_or("dataset", "software")

# We fetch the data using the `ds_and_soft` string passed to the "type" filter
ds_or_soft_2026_15_01 <- oag_query(
  entity = "research-products",
  type = ds_and_soft,
  options = oag_options(page = 1, pageSize = 1)
) |>
  oag_request()

# Print the number of records available in the response's header
ds_or_soft_2026_15_01[["header"]][["numFound"]]
```

Since there are only four types of research products, we can also negate the two others to get the same results.

```{r}
# Concatenate the strings "publication" and "other" with "OR" and negate it
not_others_or_pub <- concat_not(concat_or("publication", "other"))

# We fetch the data using the `not_others_or_pub` string passed to the "type"
# filter.
not_pub_or_other_2026_15_01 <- oag_query(
  entity = "research-products",
  type = ds_and_soft,
  options = oag_options(page = 1, pageSize = 1)
) |>
  oag_request()

# Print the number of records available in the response's header
not_pub_or_other_2026_15_01[["header"]][["numFound"]]
```
